---
title: "Re-analysis of GEO GSE54825: transcriptional response of engineered *Saccharomyces cerevisiae* to cellobiose versus glucose"
subtitle: "A reproducible pipeline with a quantitative comparison to the original publication"
author: "ApaTitee"
date: today
date-format: "YYYY-MM-DD"
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    fig-align: center
    code-fold: true
    theme: cosmo
  typst:
    toc: true
    number-sections: true
    margin:
      x: 2.2cm
      y: 2.2cm
    mainfont: "Noto Sans SC"
bibliography: refs.bib
reference-section-title: "References"
---

```{r doc-setup}
#| echo: false
#| message: false
#| warning: false
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

{{< include _setup.qmd >}}

# Background {#sec-background}

*Saccharomyces cerevisiae* cannot metabolise cellobiose. Engineering a heterologous cellobiose
utilisation pathway — a cellodextrin transporter (cdt-1) and an intracellular β-glucosidase
(gh1-1) from *Neurospora crassa* — enables growth on this sugar, but fermentation remains slower
than on glucose [@Galazka2010]. The original study of the data set re-analysed here
(GEO [GSE54825](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE54825); [@Lin2014])
profiled the transcriptome of the engineered strain on cellobiose versus glucose under anaerobic
conditions and reported 519 differentially expressed genes, mitochondrial activation, reduced
amino-acid and thiamine biosynthesis, and two engineering leads (SUT1 overexpression,
HAP4 deletion).

That study was analysed with a commercial platform (CLC Genomics Workbench 6.5), a private
reference assembly, and a statistical framework (Baggerley's test) that is no longer standard.
This report re-analyses the same raw reads with a modern, fully scripted pipeline and asks a
question the original could not answer: **how much of the original result is reproducible, and
what is not assessable at all?**

# Research question {#sec-question}

What is the global transcriptional difference between cellobiose- and glucose-grown engineered
*S. cerevisiae* under anaerobic exponential growth; which biological processes are involved; and
which observations from the original study can be quantitatively reproduced with an independent
pipeline?

# Dataset {#sec-dataset}

The seven input files (paper and supplements of both studies) are archived read-only with md5
checksums; the sequencing data are public and were verified against the md5 values published by
ENA.

{{< include _d_dataset.qmd >}}

# Experimental design {#sec-design}

Two conditions × three biological replicates, a single batch, anaerobic exponential phase
(80 g/L starting sugar, harvested at ≈50 g/L residual sugar). The original study used an
unpaired two-group comparison; replicate numbering (`JCYL001B/001D`, …) hints at pairing but this
cannot be verified from public metadata. The primary model is therefore unpaired (`~ condition`)
and a paired model (`~ replicate + condition`) is run as a sensitivity analysis
(see the *Differential expression* section). No spike-in controls and no independent batch are available.

# Methods {#sec-methods}

```{r tab-methods}
#| echo: false
m <- data.frame(
  step = c("Raw data", "Read QC", "Reference (host)", "Reference (transgenes)",
           "Quantification", "Gene-level matrix", "Differential expression",
           "Multiple testing", "Functional enrichment", "Report"),
  tool_and_version = c(
    "ENA FTP download; md5 verified against ENA-published values",
    "FastQC 0.12.1; fastp 1.3.6 (light trimming, length >= 36 bp); MultiQC 1.35",
    "RefSeq GCF_000146045.2 (S288C R64-1-1, 16 chromosomes + mitochondrion); transcripts extracted with gffread 0.12.9; annotation from the SGD GAF (release R64-5-1)",
    "Host-vector-insert sequences adjudicated from the reads themselves (see the *Quantification* section)",
    "Salmon 2.7.0, decoy-aware index (k = 31; k = 25 as sensitivity), --libType A --gcBias --seqBias, fragment-length prior set to the observed read length (49 bp)",
    "tximport 1.34.0, countsFromAbundance = lengthScaledTPM",
    "DESeq2 1.46.0; primary ~ condition, sensitivity ~ replicate + condition; LFC shrinkage (apeglm) for display only",
    "Benjamini-Hochberg",
    "clusterProfiler 4.14.0 (ORA + GSEA) with self-built TERM2GENE tables; KEGG (sce) cached",
    "Quarto 1.9.38 (HTML and PDF via Typst)"),
  stringsAsFactors = FALSE)
kable(m, booktabs = TRUE, longtable = TRUE, col.names = c("Step", "Tool, version and settings"))
```

**Thresholds.** Primary DEGs are BH-adjusted p < 0.05 and |log2FC| > 1. A relaxed (p < 0.1) and a
stringent tier (p < 0.01 and |log2FC| > 1.5) are also reported, plus a tier matched to the
original study (FDR ≤ 0.001 and |log2FC| ≥ 1). **All thresholding uses unshrunken (MLE) log2FC
values**; apeglm-shrunken values are used only for ranking and plots. The enrichment universe is
the set of expressed genes (count ≥ 10 in at least three samples): `r` `r` 6049 genes.

# Quality control {#sec-qc}

The 2014 data are single-end 50 bp reads from a Genome Analyzer II. FastQC reported
Sanger/Illumina 1.9 encoding, so no Phred64 handling was required. Light trimming retained 95.1%
of 238.6 million reads and raised the Q30 fraction from 0.88–0.93 to 0.93–0.96; adapter content
is negligible (≈150 reads per sample after trimming, 0.01%).

*Read fate.* Reads were attributed in three tiers: (i) end-to-end alignment to the host genome
plus transgenes (94.5–95.5%), (ii) local-mode re-alignment of the remainder (recovering
19.9–22.8% of the unmapped fraction, attributable to the strict end-to-end mode and to read
errors in these older data), and (iii) a diagnostic reference containing the pRS426 backbone, the
native 2 µm plasmid and TruSeq adapters (a further 35.6–40.4% of the residual). A small,
quantified fraction (2.2–3.0% of the library) remains unexplained; candidate explanations are
listed in the *Limitations* section. Multi-mapping is 23.0–24.9%, driven by Ty elements and the
rDNA repeat, and the rDNA locus is the most deeply covered region of the reference
(RDN25-1, mean depth 23×).

*Plasmid copy number.* Reads from the engineered plasmid backbone are 1.65× more abundant on
cellobiose (10,385 → 17,188 reads per million), whereas reads from the native 2 µm plasmid are
unchanged (3,358 → 3,012). This is independent evidence that any copy-number effect is specific
to the engineered plasmid.

# Quantification {#sec-quant}

The host reference and annotation were frozen (RefSeq `GCF_000146045.2`, SGD annotation release
R64-5-1, all files md5-recorded). Two properties of this data set required specific handling.

**The transgenes are not what the literature suggests.** The RNA-seq samples carry
pRS426-BT, whose inserts derive from *N. crassa* cDNA. Reads-derived consensus sequences match the
public CDS of cdt-1 (NCU00801) and gh1-1 (NCU00130) base-for-base (0 differences). The
codon-optimised `gh1-1a` published in the supplement (Figure S8) is **absent** from the samples
(2 reads in 6 million) and differs from the native CDS at 25% of positions: using it as a
reference would have silently destroyed transgene quantification. The C-terminal tag is a
**superfolder GFP**: the consensus differs from the humanised EGFP CDS at only 7 nucleotides,
and its translation carries all six superfolder substitutions (S30R/Y39N/N105T/Y145F/I171V/A206V)
on an EGFP backbone. The cdt-1–tag junction was read directly from soft-clipped reads:
`…GTTGCT` + `ATCGAT` (ClaI) + `GGTAGTGGTAGT` (Gly-Ser-Gly-Ser) + `GTGAGC…`, consistent with the
cloning strategy described in the earlier work [@Galazka2010]. Because part of the transgene
reference is derived from the sample reads, transgene results are reported **as ratios only**.

**Mitochondrially encoded genes must be measured, not assumed.** The RefSeq RNA FASTA contains no
mitochondrial protein-coding mRNA, so transcripts were extracted from the GFF with gffread
(6480 host transcripts, including 19 mitochondrial mRNAs) plus the two transgenes.

Salmon with a decoy-aware index mapped 80.0–83.5% of reads (k = 31); library type was inferred as
unstranded (**U**) in all twelve runs, consistent with TruSeq v1 chemistry. Quantification was
validated in three independent ways: (i) TPM versus the original RPKM correlates with
Spearman = 0.880 over 5796 genes; (ii) adding or removing the genome decoy changes TPMs by a
median of 0.027 log2 units (17 of 5960 genes exceed 1); and (iii) two calibration genes reproduce
the original values almost exactly (URA3 ratio 1.364 versus 1.354; PGK1 ratio 0.749 versus 0.745).

> **Reported `EffectiveLength` values are anomalous in Salmon 2.7.0 (Rust port)**: host
> transcripts receive a median effective length of 0.465 × length, and the transgene entries
> receive values above 10 × length. This does not affect the conclusions above; the checks in
> (i)–(iii) and the decoy comparison were performed specifically to bound the risk, and all
> transgene statements are based on ratios.

# Sample-level analysis {#sec-sample}

```

{{< include _d_samplelevel.qmd >}}

Conditions separate on PC1 (93.4% of variance) and replicates within a condition are extremely
concordant (Spearman ρ 0.994–0.998) compared with the between-condition range (0.886–0.895),
supporting H0.

# Differential expression {#sec-de}

{{< include _d_de.qmd >}}

The primary model yields **644 DEGs** (346 up, 298 down); the tier matched to the original
thresholds yields **632** (341 up, 291 down). The paired sensitivity model yields 643 DEGs and
agrees with the primary model almost perfectly (Jaccard 0.986; 100% directional concordance;
Pearson correlation of log2FC = 1.0000), so the conclusions do not depend on the pairing
assumption. Model diagnostics show no systematic anti-conservatism and only one gene above the
Cook's distance cutoff.

# Functional enrichment {#sec-enrichment}

{{< include _d_enrichment.qmd >}}

The original study used FunSpec with Bonferroni correction (p 0.01); this re-analysis uses
clusterProfiler with Benjamini–Hochberg correction and a self-built GO slim mapping (SGD
`go_slim_mapping.tab`) as well as full GO annotations propagated to ancestor terms via the true
path rule. The direction and identity of the enriched processes reproduce the original
conclusions: upregulated genes are dominated by the respiratory chain, oxidative
phosphorylation, ATP synthesis, mitochondrial and TCA-cycle terms, whereas downregulated genes
are dominated by amino-acid biosynthesis (methionine, cysteine, arginine, histidine) and
**thiamine (vitamin B1) biosynthesis**, with the same signal in KEGG (oxidative phosphorylation,
biosynthesis of amino acids, cysteine/methionine and histidine metabolism).

One observation not reported originally: the strongest negative GSEA signal involves Ty
retrotransposon-related terms. Because these sequences are heavily multi-mapping, this finding is
reported with caution and is not used for any biological claim.

# Biological interpretation {#sec-interpretation}

{{< include _d_scorecard.qmd >}}

The nine pre-registered hypotheses were scored with set-level statistics (Wilcoxon signed-rank
against zero, Fisher's exact test for differential-expression enrichment, and the number of DEGs
inside each set). Every hypothesis is supported:

- **H1 (mitochondrial activation)** replicates across five independent gene sets
  (respiratory chain complex +1.91, ATP synthesis +1.79 with 15 of 16 genes differentially
  expressed, oxidative phosphorylation +1.79, electron transport chain +1.55, TCA cycle +0.86).
  Mitochondrially encoded genes are excluded from this test (see *Limitations*).
- **H2 (amino-acid and thiamine biosynthesis down)** replicates in all six subsets, including
  thiamine biosynthesis (−1.22, 7 of 16 genes differentially expressed, Fisher p = 2.7 × 10⁻³).
- **H3 (glucose signalling only partially activated)** is supported in the sense the original
  reported: the PKA set shifts only slightly (+0.45) and the Rgt1 and Snf1–Mig1 sets show no
  systematic shift.
- **H4 (hexose-transporter rearrangement)** has no net directional shift (mean log2FC −0.14) but
  six of eleven transporters are differentially expressed: a family-level rearrangement rather
  than a uniform response.
- **H5–H8** replicate, including the engineering leads: SUT1 +1.77, HAP4 +2.61 and DAL80 +1.22.
- **H9 (growth-rate/glucose-repression confounder)** — the ribosomal-protein gene set (132 genes)
  shifts by +0.34 (p = 1.4 × 10⁻²¹). A confounder therefore exists and is measurable; its
  direction suggests relief of glucose repression rather than a slow-growth effect, because
  ribosomal-protein genes are typically *up* at higher growth rates in yeast.

# Comparison with the original study {#sec-comparison}

{{< include _d_comparison.qmd >}}

Agreement is high at every level that can be compared:

| Level | Result |
| --- | --- |
| Per-gene log2FC (5771 genes) | Spearman **0.973**, Pearson 0.840 |
| DEG overlap (632 vs the original 519) | **496 shared**, Jaccard 0.757, hypergeometric p ≈ 0 |
| Direction concordance in the overlap | **100%** |
| Count reconciliation | original 519 = 244 up + 256 down + 18 unmappable + *gh1-1* (no GO annotation) |
| Transgene fold changes (ratio-based) | 1.89–2.01 (cdt-1EGFP) and 2.22–2.32 (gh1-1) versus 1.87 and 2.16 originally |
| Plasmid copy-number proxy (URA3) | 1.364 versus 1.354 originally |

Absolute RPKM/TPM values cannot match the original: CLC's internal multi-mapping handling is not
reproducible outside CLC (only 76.3% of reads were uniquely mapped there), and percentages such
as 92.3/83.4/76.3% are software-specific quantities. What is reproducible is the gene-level trend,
and the trend is reproduced almost exactly.

# Limitations {#sec-limitations}

- **Enrichment framework differs** from the original (clusterProfiler + BH versus FunSpec +
  Bonferroni), so enrichment p-values are not directly comparable; only the identity and
  direction of the enriched processes are.
- **Mitochondrially encoded genes are not assessable**: poly-A selection captures them poorly
  (5–18 chrM reads per million; none of the 19 mitochondrial genes reaches 1× coverage), which is
  expected because budding-yeast mitochondrial mRNAs lack the long poly(A) tails used for capture
  [@Chang2012]. H1 is therefore tested on nuclear-encoded mitochondrial genes only.
- **Transgene abundance is circular to a degree**: the tag and junction sequences are derived from
  the sample reads. Ratios are robust to this (the same reference bias applies to both
  conditions), absolute TPM is not.
- **Plasmid copy number varies between conditions** (URA3 1.364×; backbone reads 1.65×). Dividing
  the transgene ratio by the proxy (1.47 for cdt-1EGFP, 1.70 for gh1-1) lowers the effect but does
  not remove it, so part of the transgene increase reflects copy number rather than transcription.
- **Terminator/promoter crosstalk** from the plasmid affects PGK1 and CYC1 minimally but cannot be
  fully separated; CYC1 is itself a mitochondrial-activation gene and changes by 2.5× in the
  original data.
- **Unmapped reads are only partly explained**: 2.2–3.0% of the library remains unattributed;
  candidates include read errors in GA II data, non-yeast contamination, regions that are
  collapsed or absent in the R64 assembly (subtelomeric/Y′ elements), and chimeric reads spanning
  junctions that exist in no reference (promoter–CDS, CDS–terminator).
- **No spike-in, no independent batch, n = 3** limit absolute quantification and variance
  estimation; surrogate-variable methods are not applicable with six samples.
- **`EffectiveLength` semantics are anomalous in Salmon 2.7.0** (see the *Quantification* section); the
  conclusions were shown to be insensitive to it, but absolute TPM values should not be
  over-interpreted.
- **Pairing could not be verified**; the paired model is reported as a sensitivity analysis only.

# Engineering implications {#sec-engineering}

The re-analysis reproduces the two engineering leads of the original study and places them in a
quantified context: SUT1 (+1.77) and HAP4 (+2.61) are among the most strongly induced transcription
factors, and DAL80 (+1.22) also qualifies. Because HAP4 is *induced* on cellobiose while its
deletion improves fermentation, its regulatory role is non-monotonic — a caution against reading
transcriptional direction as a direct engineering instruction. The yeast mitochondrial and
amino-acid/thiamine pathways together account for the bulk of the response, which suggests that
engineering strategies aimed at cofactor and precursor supply (rather than at sugar transport
alone) are the most promising.

# Reproducibility {#sec-reproducibility}

Everything in this report is generated by numbered scripts in `scripts/`:

```bash
conda env create -f environment.yml && conda activate yeast-cellobiose-rnaseq
bash run_all.sh          # idempotent; skips completed steps
```

| Artefact | Location |
| --- | --- |
| Raw-data checksums (self-computed and ENA-published) | `checksums/` |
| Reference and annotation versions | `refs/VERSIONS.md` |
| Provenance of every non-public sequence | `refs/custom/PROVENANCE.md` |
| Sequence adjudication and junction evidence | `refs/custom/{sequence_adjudication.tsv,junctions.tsv}` |
| Environment lock (conda-explicit + R package snapshot) | `env/` |
| Parameter and scope decisions | `docs/decisions/` |
| Stage acceptance records | `docs/acceptance/` |

The R environment is locked with conda rather than renv because renv project activation masks the
conda library in this setup (`docs/decisions/r_environment.md`).

# Conclusion {#sec-conclusion}

Re-analysing GSE54825 with an independent, fully scripted pipeline reproduces the original study
almost completely at the level at which comparison is meaningful: 97.3% rank agreement of per-gene
fold changes, 496 of the original 519 DEGs recovered, 100% directional concordance, and the same
enriched processes. Three results go beyond the original: the transgene reference sequences were
reconstructed and validated from the data (revealing that the published codon-optimised gh1-1a is
not the allele in these samples), the plasmid copy-number contribution to the transgene signal was
quantified, and the growth-rate/glucose-repression confounder was measured rather than assumed.
The limitations that cannot be removed — mitochondrially encoded genes, absolute transgene
abundance, and CLC-specific mapping statistics — are stated explicitly rather than papered over.

# References {.unnumbered}

::: {#refs}
:::
